Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest

### 2019-12-17 ms/response_to_reviews.tex

We thank the editor and reviewer for their helpful comments and suggestions. We
have itemized each comment below and followed each with our reply and a list of
changes we made in response to the comment.

{\bf Responses to Senior Editor's comments:}

\begin{quote}
I support the AE's recommendation to reject but allow resubmission. However, the bar for a resubmission will be high. My feeling is that your study confirms the essential findings from your 2012 study published in Journal of Ecology with the addition that your methodology of drying down adds some realism. Indeed, the confirmatory nature of this study was pointed out by reviewer 2 in their point 6. In addition to the points raised by the AE and reviewers, I will be looking in any resubmission for more evidence that this study advances your earlier work (apart from your last paragraph, your discussion is largely an explanation of your results with little context from other studies), that the findings have real ecological relevance (i.e., does it matter to the mixed-conifer communities?) and how is the new dry down treatment related to dry down conditions in litter mixtures in the field. If you can adequately address all of these concerns then I'll be happy to consider a resubmission.
\end{quote}

We agree that our prior Discussion was weak and have extensively rewritten it.
We argue that this work is not merely confirmatory. Our revised Introduction
argues for the ecological significance of dry down rates. The importance of
fuel moisture is not under debate. Furthermore, fine scale heterogeneity in
fire behavior has been statistically associated wwith variation in fuel
moisture (Hille and Stephens 2005) and fire history can vary over short spatial
scales as a consequence of species composition (Stephens 2001). Although the
importance of dry-down rates is less well-established, prescribed burns in
Sierra Nevada mixed conifer forest demonstrate that fuel moisture likely drives
differences in fire behavior depending on the timing of fire throughout the
season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille and Stephens
2005, Knapp et al 2006). These differences in fire behavior in turn control
tree mortality ((Stephens et all 2002, Schwilk et al 2006). Despite this, we
know of only one other study specifically investigating differential drying
behavior (Kreye et al 2013).

TODO
Add context also suggested by reviewer 1.

@@

TODO

{\bf Responses to Associate Editor's comments:}

\begin{quote}
First, apologies to the authors for the long delay in the review process. Because of the rather contrasting assessments of the first two reviewers, I was waiting for the third reviewer to produce his/her review, but unfortunately we waited in vain. So I have to make a recommendation based on these two contrasting assessments plus my own judgement. On the one hand, in agreement with reviewer 1, I can see the novelty of this study. So far non-additivity in litter mixture flammability has been studied either at air-dry litter moisture or at fixed levels of litter moisture (e.g. Blauw et al 2015) and I do applaud the approach of bringing more field realism into these studies by allowing mixtures to dry out at different rates from expected values based on their single components. Having said that, there are also major problems with the current manuscript, as reported especially by reviewer 2 (and some additional ones by myself, see below). Besides these major problems there is a rather high degree of inaccuracy in the text.  Therefore, given the high rejection rates of often even good manuscripts to J. Ecol., I have to make a negative recommendation for this manuscript. However, given the novelty and principally nice design of the experiment, I would allow the authors to submit a strongly revised new manuscript, addressing all major and minor comments by the reviewers and myself, as a new submission. (At least one of the reviewers to check the revision would be new.) In addition to the points raised by the reviewers, I like to add some of my own:
\end{quote}

Thank you. Several comments converged on the absolutely correct criticism that our Discussion did not engage adequately with the literature. We have made substantial changes to the Introduction and the Discussion adding more context and citing more of the recent literature. We reorganized the Introduction and now do not rush the reader so quickly through the ideas. We have expanded our justification for the study as argued for by the Senior Editor.

@@

\begin{quote}
The authors mention leaf size explicitly as an important parameter in non-additivity (with some references including their own 2012 paper) but they do not include it in the current study. Especially at low moisture, the small individual needles of Abies should cause low ventilation and a high likelihood of negative non-additivity on flammability, as was shown by Zhao et al 2016 Ecol. Evol. Figure 4). In the Methods, as observed also by reviewer 2, there is very little detail on the extent to which litter of the respective species was picked up as individual leaves, branchlets without leaves versus branchlets with leaves still attached. This is really important, also because positive versus negative non-additivity of leaf-twig mixtures have been reported recently (Zhao et al. 2019 Plant and Soil). I feel that data and/or photos indicating the size and shape of the different litter types (and if possible variability in these) are important to add and to be used in the interpretation of the results.
\end{quote}

We have added information to the methods explaining the litter collection in more detail. All material was naturally dropped litter which could include fine twigs and leaves. Litter in these two \emph{Abies} species is a combination of branchlets with still attached needles and individual needles that have detached from the branchlet.  We now cite the Zhao et al (2019) results when discussing the mixture results.

Yes, when deciding how to limit the number of tests at the comparitive scale (species as data points), we decided to not include size, because we believed its influence on flammability is primarily through litter packing and was therefore captured by litterbed density. We have added a figure (SI fig. 1) to the SI materials showing litter particle size (the containing prism volumes) distributions for each of these species. These data were collected from the stored litter. Therefore, the data show the particle size distributions of the material that was burned. We did not add these data to the main figures as they are not part of any statistical test or main conclusions, but we are happy to do so if the editor suggests.

@@

\begin{quote}
In the Methods, it was unclear how the various treatments and replicates were grouped over time, given that temperature and humidity in the fire facility could vary. Were there blocks in time with one replicate for each of the treatments (in random burning order) within a block? Or was there a different statistical design in time? 
\end{quote}

Guided by the drying rates measured in the initial trials, we staggered starting saturation times across species. This was aimed at allowing a completely blocked design. However, because some ignition attempts were unsuccessful, some days lacked the full complement of species while others, later in the process, sometimes had more than one replicate per species burned. We've added this information to the text [Lines X-X].

@@

\begin{quote}
Line 196 (see also reviewer 2): how can we assume that the quality of the stored litter is the same as that used in the previous study 3 years earlier, i.e. can these two datasets be safely combined?
\end{quote}

We have added detail as requested by reviewer 2. See lines X-X. All litter was
initially oven dried in large paper bags. The litter was then stored in the lab
at low humidity for several weeks on open racks. The litter was then bagged and
sealed and stored within garbage bags inside cardboard boxes in a lab storage
area. We saw no signs of mold nor ongoing decomposition. We minimized handling
of the bags to avoid mechanical damage and changes to litter particle size.

@@

\begin{quote}
Line 365 (see also reviewer 1): on the other hand Cornwell et al. 2015 New Phyt reported mostly high flammability in a range of Cupressaceae.
\end{quote}

This is a good point and another place where we did not discuss the literature fully. We
have now added this to the discussion and made the conjecture that these
differing patterns may result from different structure across the
Cupressaceae with Australian \emph{Callitris} and the two Californian species
packing more tightly and resulting in less flammable litter than the Asian
Cupressaceae in Cornwell et al (2015).

@@

The study has been well thought through and well executed, and the ms is generally well written. Stylistically, the main problem is that commas are missing in many sentences. This might sound trivial, but will really help the reader.
\end{quote}

We thank the reviewer for their comments and apologize for the failure in
proofreading. The reviewer should not have had to conduct copy editing and we
sincerely apologize for that. We have taken heed of the reviewers' comments and
fixed all specific grammar errors and missing symbols. For brevity, we've
omitted responses to the very simple suggested grammar fixes and simply made all
requested changes. We respond to other points individually.

\begin{quote}
line 72 - the comment about natural lightning ignitions comes from nowhere, and is a  bit of a jolt for the reader - rephrase slightly, e.g. many fires result from lightning ignitions, which are concurrent with ....
\end{quote}

This portion was poorly written and this was also mentioned by Reviewer 2. We have rewritten this in response to reviewer 2's comment that most fires result from dry lightning strikes across the west. See our responses to Reviewer 2, below.

@@

Yes. We have made edits to

TODO. Check wording.

\begin{quote}
lines 348-349 - logically, this does not necessarily follow - factors such as large leaves or high SLA could independently affect both absorptive capacity and flammability - although fuel moisture does of course have a strong effect 
\end{quote}

Yes, we think that our logic must not have been clear here as we don't see the contradiction. We  have rewritten these sentences and we hope they are no longer confusing. See lines X-X.

@@

\begin{quote}
line 403 - change to 'to the completely dry litter mixtures studied in previous work'
\end{quote}

Done. See line XX.

\begin{quote}
Figures 7 and 8 - is there a reason that you don't present the information in the same way as in Figure 6, if your main point is the discrepancy between the mixtures and the averaged individual values?
\end{quote}

We agree this does lack consistency and an early draft used the same concise format of Fig 6 for all three figures. However, we decided that having time on the x-axis was worth the space and allows the reader more insight into the flammability effects of drying over time. Reviewer 2 pointed out some exceptions to the overall statistical pattern and those exceptions would not have been easily visible without an explicit time axis. There is a trade-off here between illustrating simply the results of the statistical model and allowing the reader to get a better idea of variation in flammability consequences of drying time.

@@

{\bf Responses to the Reviewer 2's comments}

\begin{quote}

My major concerns are:
1. One motivation for the study is not valid. I do not completely agree with the authors that natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event. The efficiency of individual lighting strikes in igniting a fire is affected by the moisture properties of forest fuel at the moment of lighting (Krawchuket et al. 2006, Ecology). Dry lighting that occurs without any accompanied rainfall nearby is most likely to cause wildfire because the fuel is dry enough to be ignited.
\end{quote}

This sentence was poorly written and oversimplified. We thank the reviewer for
pointing this out and we have rewritten this section. However, our badly made
point was: lighting ignitions can occur during the same general period as
precipitation; therefore, drying rates are important. As the reviewer points
out, dry lightning strikes that find dry fuel are more likely to lead to fires
than are strikes that find damp fuels (Abatzoglou et al 2016). The fact that
the total number lightning strikes does not well predict burned area and that
ignitions occur often from dry strikes argues for the importance of fuel
moisture. Although synoptic weather conditions that create dry and wet storms
differ, the potential for precipitation and lightning during the same overall
period of thunderstorm activity often exists (eg wet and dry convection days
can occur under overlapping conditions, Rorig and Ferguson 1999). We've
rewritten this section, see lines XX-XX.

@@

\begin{quote}
2. The research question of the research is not explicitly stated both in the abstract and in the introduction. In the end of the introduction, the authors focus on specific data-collection goals, while leave the reader to figure out what the research question is. 
\end{quote}

We have edited the abstract to make the research questions clearer. Yes, our
first submission jumped from stating current reserach gaps to specific methods
too quickly. See also our response to the Senior Editor, above.

\begin{quote}
3. The authors concluded that positive non-additivity appeared to be the rule in litter-driven fire and this non-additivity became more positive with fuel drying. But if we look closely at figure 8, non-additivity was negative within 50 hours since wetting, non-additivity was both positive and negative at about 75 hours since wetting, and non-additivity was positive at about 100 hours since wetting. The high variability of non-additive effects at about 75 hours since wetting might have important ecological effects on fire ignition and behavior, and are important information cannot be ignored.
\end{quote}

Yes, this is a good point and we have added some discussion of this. We tried to limit our statements to those supported by the statistical tests (eg, slope of nonaddivity with drying was greater than 1), and therefore focused on the positive effect of drying and the overall trend for positive nonadditivity. This, of course, obscures some details which are worth mentioning and we have edited this section in response to the reviewer's criticism. We agree that the high variability in nonadditivity across mixtures deserves highlighting and that is the reason we presented the actual predictions and departures from prediction across time. Please see our changes on lines X-X.

@@

\begin{quote}
The abstract can be shorter. It exceeds 350 words limit. 
\end{quote}

We have editing the abstract to more clearly state the research questions reduce its length.

\begin{quote}
Line 50: add reference to the first sentence.
\end{quote}

We thought this was an established enough fact to not need citation but have added one.

@@

\begin{quote}
Line 88: “similarly”?
\end{quote}

We were not sure what the reviewer's criticism was here but have rewritten this system to be more detailed and precise.

\begin{quote}
Line 95: “absorption capacity” and “rate of moisture loss” are not correct descriptions of processes. Better say “absorption and retention”.
\end{quote}

Good point. We fixed that here and in another sentence elsewhere. Furthermore, we've edited to be more precise about what we mean by ``absorption'' because we are referring to a litter bed and we include adsorption to litter surfaces in that definition. We had originally tried to avoid the chemical terms because we are observing a higher-level phenomenon.

@@

\begin{quote}
Lines 103-105: leaf size is not leaf litter trait.
\end{quote}

We have rewritten this. See line X.

\begin{quote}
Line 107: “fuel-beds composed of larger leaves with lower moisture retainability”?
\end{quote}

We edited to ``composed of larger leaves or leaves with lower moisture retainability''

### 2019-12-18 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Rewrite conclusions, make edits
based on first proof read.

\nwords{7600}
\ntables{4}
\nfig{8}
\nref{39}

@@

      \item Moisture content is a strong determinant of forest fuel
    flammability, however, little is known about the mechanisms by which
    species-specific litter traits influence litter moisture absorption and
    retention. Characterizing how species’ leaf litter retain moisture through
    time and how fuel from multiple species interacts are critical to the
    assessment of fuel’s availability to burn and key steps to adding species
    or taxon specific information to fire behavior prediction.

      \item To fill these gaps in knowledge, we determined
    moisture absorption and retention characteristics of the litter from
    multiple tree species, investigated what litter traits influenced these
    processes, and tested for nonadditive moisture dynamics and nonadditive
    flammability in multi-species mixtures. The litter from eight tree species of a
    mixed-conifer forest was saturated and allowed to dry to determine moisture
    absorption capacity and drying rates. Burn trials were performed in moist
    litter beds of single species and of mixtures to establish flammability
    response to the dry-down process.

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    result in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.
    Litter mixtures exhibited consistent nonadditivity in flame spread rate and
    in fuel consumption. This nonadditivity became more positive with fuel
    drying.

      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
    and retention can change flammability rankings under moist compared with dry fuels.
    Litter mixtures generally burn with fire behavior more similar to that of
    the most flammable constituent species and this effect increases as fuels
    dry. [TODO more]

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond very readily
to environmental changes \citep{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels
varies with climate, with seasons, and with short term weather within a season.
Therefore, the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season \citep{Nelson-2001}.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with rates of heat release and total burned area and with heat
transfer to soils or tissues predicting soil impacts of fire and the mortality
of better insulated plant tissues. A shortcoming in our current knowledge is
our integration of specific species and plant traits into fuel hazard
assessment, fire modeling, and fire effects prediction. Developing quantitative
yet robust and general ways to integrate actual plant communities into fuel
assessment will require understanding the magnitude of flammability variability
across species in space and time and parameterizing based on plant
characteristics. [TODO: get opinion on final sentences]

Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015} and fire regime can vary of small spatial scales in response to community composition and species-specific flammability \citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter packing, measured as litter-bed density or
packing ratio, is an important control ignition and spread in models of surface
fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
Multiple studies have demonstrated that litter bed density controls the
flammability of dry litter fuels in seasonally dry climates where higher litter
density reduces flame spread rate. Across these studies, there is increasing
consensus that larger less densely packed litter beds increase oxygen
availability and burn with faster spread rates \citep{Scarff+Westoby-2006,
  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
by litter particle size and shape which is in turn influenced by leaf size, how
leaves or branches are abscised, and subsequent decomposition
\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
  Prior+Murphy+etal-2018, Zhao+vanLogtestijn+etal-2019}.

These species-specific effects are increasingly well established. However, how
moisture dynamics modulate species specific flammability is poorly known. Species
specific effects on litter flammability are recorded for the forests of the
southeastern US \citep{Nowacki+Abrams-2008} where species-specific litter
drying rates play a role in flammability feedbacks
\citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual
species traits have been recorded for western US forests as well: in mixed
conifer forest in California, where the fire regime is characterized by surface
fires fueled by plant litter, overstory tree composition drives variation in
local fire behavior \citep{Schwilk+Caprio-2011} as a result of differential
litter flammability across tree species and as a result of synergistic
interactions in multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}.
These flammability feedbacks may help explain historical fire frequency
patterns \citep{Schwilk+Caprio-2011}, but current estimates of litter
flammability in mixed conifer forests are based largely on burning dry litter.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Drying rates following snowmelt
or following rainfall during the fire season will influence the length of the
fire season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
across time may contribute to the often high spatial heterogeneity of surface
fires \citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and
to variation in fire regime across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting (including absorption and adsorption) and drying rate (desorbtion) \citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated. Differences in moisture content over time since
saturation (drying rate) can indicate different abilities to retain moisture. Litter particle
geometry may have direct effects on maximum moisture absorption through
area-mass relationships and we also expect effects of leaf morphology,
especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption,
moisture retention should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore effects fire behavior in two
ways: directly by its effect on oxygen availability for combustion, and
indirectly by its effect on dry-down rate. Both these pathways are affected by
leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
Fuel-beds composed of larger leaves or leaves with lower moisture retainability should
favor drying, thus increasing the probability of ignition and fire spread.

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible that
similar nonadditive effects may influence litter moisture content directly and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}.

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
estimates based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water retention and rate of moisture loss and examined
how these vary in relation to specific leaf area and litter packing density.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. We tested if the nonadditive effects on reported in
previous studies occur under higher moisture levels, and whether fuel moisture
effects strengthen or moderate any nonadditivity.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
by changing the horizontal dimensions of the basket. Weight was measured by
using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
OH). Samples were immersed in plastic storage bins filled with water for 24
hours to fully saturate them. After saturation, the baskets were removed from
the water, allowed to drain for 3 minutes, and samples weighed to measure
maximum moisture content. The samples were then allowed to dry in a controlled
environment kept at $21^\circ$C temperature and 30\% relative humidity,
assessed via an iButton sensor (Maxim Integrated) placed over the samples.
Litter baskets were arranged in a completely randomized design. During the
dry-down process three subsamples of 5--10\,g were taken at 24 hour intervals
and weighed. These subsamples were then oven dried for 24 hours at $100^\circ$C
and re-weighed to assess moisture content. This was repeated until the final
sample from a basket was less than 10\% moisture. This experimental setup was designed to accelerate drying compared with field conditions because the baskets were suspended in dry air 2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the closest weather stations to the collection sites (Grant Grove and Lodegpole stations, NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).

>>>>

\nwords{8200}
\ntables{4}
\nfig{8}
\nref{51}

@@

      \item Moisture content is a strong determinant of forest fuel
    flammability, however, little is known about the mechanisms by which
    species-specific litter traits influence litter moisture absorption and
    retention. Characterizing how litter fuels retain moisture through time and
    how fuels from multiple species interact are key steps to adding species or
    taxon specific information to fire behavior prediction.

      \item To fill these gaps in knowledge, we determined moisture absorption
    and retention characteristics of the litter from multiple tree species,
    investigated what litter traits influenced these processes, and tested for
    nonadditive moisture dynamics and nonadditive flammability in multi-species
    mixtures. The litter from eight tree species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in litter beds of single species
    and of mixtures to establish flammability response to the dry-down process.

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    result in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.
    Litter mixtures exhibited nonadditivity in flame spread rate and
    in fuel consumption. This nonadditivity became more positive with fuel
    drying.

      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
    and retention can change flammability rankings under moist compared with
    dry fuels because traits that increase flame spread rate increase water
    absorption. Litter mixtures generally burn with fire behavior more similar
    to that of the most flammable constituent species and this effect increases
    as fuels dry.

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond very readily
to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
with climate, with seasons, and with short term weather within a season.
Therefore, both the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season \citep{Nelson-2001}.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with rates of heat release and area burned, and with heat
transfer predicting soil impacts of fire and the mortality. A shortcoming in
our current knowledge is our integration of specific species and plant traits
into fuel hazard assessment, fire modeling, and fire effects prediction.
Developing quantitative yet robust and general ways to integrate actual plant
communities into fuel assessment will require understanding the magnitude of
flammability variability across species in space and time and parameterizing
based on plant characteristics.

Species differ in their litter flammability \citep{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015} and fire regime can vary over fine spatial scales
in response to community composition and species-specific flammability
\citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter
packing, measured as litter-bed density or packing ratio, is an important
control on ignition and spread in models of surface fire \citep{Weber-1991,
  Balbi+Santoni+etal-1999, Morvan+Larini-2001}. Multiple studies have
demonstrated that litter bed density controls the flammability of dry litter
fuels in seasonally dry climates where higher litter density reduces flame
spread rate. Across these studies, there is increasing consensus that larger,
less densely packed litter beds increase oxygen availability and burn with
faster spread rates \citep{Scarff+Westoby-2006, Schwilk+Caprio-2011,
  Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015, Varner+Kane+etal-2015,
  Kreye+Varner+etal-2018}. The litter packing is driven by litter particle size
and shape, which is in turn influenced by leaf size, how leaves or branches are
abscised, and subsequent decomposition \citep{Fonda+Belanger+etal-1998,
  Stephens+Finney+etal-2004, Scarff+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015, Zhao+Cornwell+etal-2016,
  Grootemaat+Wright+etal-2017, Prior+Murphy+etal-2018,
  Zhao+vanLogtestijn+etal-2019}.

These species-specific flammability effects are increasingly well established.
However, how moisture dynamics modulate species specific flammability is poorly
known. Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Drying rates following snowmelt
or following rainfall during the fire season will influence the length of the
fire season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
across time may contribute to the often high spatial heterogeneity of surface
fires \citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and
to variation in fire regime across landscapes \citep{Stephens-2001}.

Fuel moisture content through time is controlled by two processes, wetting
(including absorption and adsorption) and drying rate (desorbtion)
\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated; hereafter we refer to this measurement as ``maximum
water absorption''. Differences in moisture content over time since saturation
(drying rate) can indicate different abilities to retain moisture. Litter
particle geometry may have direct effects on maximum moisture absorption
through surface area affects on adsoprtion and we also expect effects of leaf
morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
rate should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing, therefore, effects fire behavior in two
ways: directly by its effect on oxygen availability for combustion, and
indirectly by its effect on drying rate. Both these pathways are affected by
leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
Fuel-beds composed of larger leaves or leaves with lower maximum moisture absorption
should favor drying, thus increasing the probability of ignition and fire
spread.

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}. In mixed conifer forest in
California, where the fire regime is characterized by surface fires fueled by
plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
estimates based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these vary in relation to specific leaf area and litter packing density.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. We tested if the nonadditive effects on reported in
previous studies occur under higher moisture levels, and whether fuel moisture
effects strengthen or moderate any nonadditivity.

@@

We produced dry-down curves to determine maximum water absorption and the
rate of drying for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
by changing the horizontal dimensions of the basket. Weight was measured by
using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
OH). Samples were immersed in plastic storage bins filled with water for 24
hours to fully saturate them. After saturation, the baskets were removed from
the water, allowed to drain for 3 minutes, and samples weighed to measure
maximum moisture content. The samples were then allowed to dry in a controlled
environment kept at $21^\circ$C temperature and 30\% relative humidity,
assessed via an iButton sensor (Maxim Integrated) placed over the samples.
Litter baskets were arranged in a completely randomized design. During the
dry-down process three subsamples of 5--10\,g were taken at 24 hour intervals
and weighed. These subsamples were then oven dried for 24 hours at $100^\circ$C
and re-weighed to assess moisture content. This was repeated until the final
sample from a basket was less than 10\% moisture. This experimental setup was
designed to accelerate drying compared with field conditions because the
baskets were suspended in air 2.8 -- 5.3$^\circ$C above the warmest month
average temperatures for the closest weather stations to the collection sites
(Grant Grove and Lodegpole stations, NOAA 1981-2010 normals,
https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
